---
title: "Checking quantitative structure with QuantFit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking quantitative structure with QuantFit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1)
```

## The question

Before treating a psychological or educational test score as a *measurement*,
it is worth asking whether the data actually support a quantitative
interpretation of the latent variable at all. The differences between people
might be quantitative (a matter of *how much*), merely ordinal (a matter of
*rank*), or only classificatory (a matter of *kind*). QuantFit provides three
complementary, empirically testable answers to that question:

1. **Conjoint cancellation checks** ask whether the response probabilities
   obey the cancellation axioms of additive conjoint measurement — the
   order-theoretic conditions that must hold if persons and items sit on a
   common additive scale.

2. **The Karabatsos omnibus test** poses a closely related question in a
   Bayesian synthetic-likelihood framework, quantifying departures from
   additivity cell by cell.

3. **Latent structure model selection** fits a hierarchy of six latent
   variable models and asks which level of structure — classificatory,
   ordinal, or quantitative — the data actually warrant.

The three approaches come at the same underlying question from different
directions; used together they give a much fuller picture than any one alone.

```{r setup}
library(QuantFit)
```

We will use a single simulated data set throughout. It follows a Rasch model,
so all three approaches *should* conclude that the structure is quantitative.

```{r simulate}
set.seed(1)
n_persons <- 800
n_items   <- 12
theta <- rnorm(n_persons)                       # person abilities
beta  <- seq(-2, 2, length.out = n_items)       # item difficulties
prob  <- plogis(outer(theta, beta, "-"))
responses <- matrix(rbinom(n_persons * n_items, 1, prob),
                    n_persons, n_items)
dim(responses)
```

## 1. Conjoint cancellation checks

The cancellation checks operate on a matrix of counts cross-classifying
*sum score* against *item*. `PrepareChecks()` builds the required success
counts `n` and attempt counts `N` from a raw 0/1 response matrix.

```{r prepare}
prep <- PrepareChecks(responses)
dim(prep$N)
```

`ConjointChecks()` then samples submatrices, runs a constrained MCMC sampler
on each, and reports the proportion of cells whose observed proportion falls
outside the posterior credibility region implied by the axiom. Low violation
rates are consistent with additive (quantitative) structure. The sampler is
the computational bottleneck, so the call below is shown but not evaluated in
this vignette:

```{r conjoint, eval = FALSE}
cc <- ConjointChecks(prep$N, prep$n, n.mat = 100, check = "double",
                     mc.cores = 1)
cc@means            # unweighted and item-count-weighted violation rates
#> $unweighted
#> [1] 0.03
#>
#> $weighted
#> [1] 0.02
```

`check = "single"`, `"double"`, and `"triple"` test progressively stronger
cancellation axioms; `HiConjointChecks()` runs the full hierarchical test over
every 3x3 submatrix embedded in each 4x4 block. The bundled `rasch1000` object
is a pre-computed result you can explore immediately:

```{r rasch1000}
data(rasch1000)
summary(rasch1000)
```

## 2. The Karabatsos omnibus test

`KaraChecks()` implements Karabatsos (2018): it fits a two-stage isotonic
model to the observed proportions, simulates synthetic data sets under a
Beta-Binomial null, and uses a synthetic-likelihood importance sampler to
obtain a posterior for each cell. Departures from additivity are summarised
by a Kullback-Leibler divergence per cell (flagged when KL > 0.01) and a
global KL. This is a Monte Carlo procedure; realistic runs use `S` in the tens
of thousands, so it too is shown rather than evaluated:

```{r kara, eval = FALSE}
kc <- KaraChecks(prep$N, prep$n, S = 30000, N_synth = 100, mc.cores = 1,
                 verbose = FALSE)
kc$global_KL      # overall departure from additivity
kc$n_violations   # number of cells with KL > 0.01
kc$KL             # per-cell KL divergence matrix
```

For Rasch-generated data the global KL is small and few cells are flagged;
strongly non-additive data (for example 2PL data with dispersed
discriminations) produce many more violations. This R/C++ implementation is
validated against the original MATLAB `ACMtest.m` to a cell-level correlation
above 0.99.

## 3. Latent structure model selection

The third approach fits six nested-in-structure latent variable models:

| Model | Structure | Interpretation |
|-------|-----------|----------------|
| `UN`  | Unconstrained latent classes | Classificatory |
| `MON` | Class monotonicity | Ordinal |
| `IIO` | Invariant item ordering | Ordinal |
| `DM`  | Double monotonicity (MON + IIO) | Ordinal |
| `LCR` | Latent class Rasch | Quantitative (discrete) |
| `RM`  | Rasch model | Quantitative (continuous) |

You can fit any single model directly:

```{r fit-un}
un <- fit_un(responses, n_classes = 3, verbose = FALSE)
un
```

### Choosing the number of classes

Every discrete model conditions on a number of latent classes, and that
choice is a modelling decision in its own right — not something to fix
arbitrarily. `select_n_classes()` makes it explicit: it fits the
unconstrained model across a range of class counts (including the
single-class baseline) and reports the enumeration so you can pick a count by
BIC before comparing structure.

```{r n-classes}
nc <- select_n_classes(responses, C_range = 1:6, seed = 1)
nc
```

The recommended two-stage workflow is therefore: choose the class count here,
then run the structural comparison at that count. `select_model_ll()` will do
both in one call if you pass a range to `n_classes` (it selects the count by
BIC internally and reports the enumeration as `n_classes_table`). Fitting many
class counts and bootstrapping is the expensive part of the analysis, so these
functions take an `mc.cores` argument that spreads the work across cores;
because every fit is seeded independently the parallel result is identical to
the serial one.

### Comparing all six by information criteria

`compare_models()` fits the whole set and ranks them. Note the caveat it is
built to expose: UN, MON, IIO, and DM all have the *same* number of
parameters, so BIC (which only rewards parsimony through parameter count)
cannot separate them — it will tend to pick UN or, for quantitative data, RM.

```{r compare}
cmp <- compare_models(responses, n_classes = 3, verbose = FALSE)
cmp
```

### Calibrated selection with the likelihood-ratio bootstrap

`select_model_ll()` addresses that limitation. It tests each constrained model
against its parent with a parametric bootstrap of the likelihood-ratio
statistic — the correct calibration when the models are nested by inequality
constraints and share a parameter count (the asymptotic null is
chi-bar-squared, not chi-squared). It walks down the hierarchy
UN -> DM -> LCR -> RM, accepting each reduction whose bootstrap test is
adequate. Even the final RM-versus-LCR step is bootstrap-calibrated: rather than
comparing the two BIC values directly, it bootstraps the BIC *difference*
against a null simulated from the fitted Rasch model (`rm_vs_lcr_test()`), so
the continuous model is preferred only when the discrete latent-class Rasch fit
is no better than Rasch data would produce by chance. Because it refits models
on bootstrap samples it is moderately expensive; here we use a small `B` for
illustration:

```{r select-ll, eval = FALSE}
sel <- select_model_ll(responses, n_classes = 3, B = 99, seed = 1)
sel
#>
#> Bootstrap LL-equivalence latent structure selection
#> ---------------------------------------------------
#> Selected model : RM
#> Interpretation : QUANTITATIVE (continuous: Rasch model)
#> Latent classes : 3
#> Method = joint   Alpha = 0.05   Bootstrap replicates per test = 99
#>
#> Decision path (bootstrap LR tests):
#>   DM vs UN   LR =    2.104   p = 0.2727   -> adequate (p > alpha)
#>   LCR vs DM  LR =    3.856   p = 0.1414   -> adequate (p > alpha)
#>   RM vs LCR  LR =    4.219   p = 0.6238   -> continuous (RM) preferred
#>
#> BIC comparison at the quantitative step:
#>   LCR  BIC = 11846.88
#>   RM   BIC = 11842.66   (selected)
```

Here the doubly-monotone reduction is adequate (DM vs UN not rejected), the
discrete Rasch reduction is adequate (LCR vs DM not rejected), and the final
bootstrap prefers the continuous Rasch model, so the data are read as
quantitative — the expected verdict for Rasch-generated data.

A purely diagnostic alternative, `select_model_constraint()`, inspects whether
the unconstrained estimates *satisfy* each ordinal constraint (within a
tolerance) rather than testing them, and is much faster:

```{r select-constraint}
sc <- select_model_constraint(responses, n_classes = 3, seed = 1)
sc$selected
sc$interpretation
```

### Standard errors

`compute_se()` returns standard errors for a fitted model, either from the
observed information (default) or by nonparametric bootstrap. For a model
whose fitted estimates sit on an active monotonicity constraint, the
Hessian-based standard errors for the affected parameters are returned as `NA`
with a warning, and the bootstrap method should be used instead.

```{r se}
se <- compute_se(un, method = "hessian")
se
```

### How much structure can be recovered?

Not every distinction is equally easy to make, and it is worth knowing where
the method is strong and where it is inherently limited. In a simulation study
(1000 respondents, three latent classes, 30 replicates per generating model),
recovery of the correct model was near-perfect for UN, DM, and RM, and strong
for MON. Two distinctions were harder, for instructive reasons:

* **Discrete versus continuous quantitative structure (LCR versus RM)
  depends on test length.** With 10 items the two are nearly interchangeable
  and LCR-generated data are often misread as RM; doubling the test to 20
  items raised LCR recovery from roughly 40% to over 70%. If the discrete /
  continuous distinction matters for your application, use a longer test
  (20 or more items) so the latent class Rasch structure can be resolved from
  the continuous Rasch model. Reassuringly, this confusion is *within* the
  quantitative interpretation, so the broad classificatory / ordinal /
  quantitative verdict is unaffected by it.

* **Invariant item ordering versus double monotonicity (IIO versus DM) is
  limited by the class structure, not the test length.** Whether an item's
  response probability increases monotonically across the ordered classes is a
  property of how many classes there are and how well separated they are; with
  only a few classes, IIO-generated data are frequently monotone by chance and
  are legitimately consistent with DM. Adding items does not resolve this - it
  requires more, or better-separated, latent classes. Both are ordinal
  interpretations, so again the scale-type conclusion is unaffected.

The practical message is that the coarse question this package is built to
answer - is the structure classificatory, ordinal, or quantitative? - is
recovered very reliably, while the finer within-family distinctions benefit
from longer tests (for LCR vs RM) or richer class structure (for IIO vs DM).

## Putting it together: `quant_fit()`

Running the three approaches by hand and reconciling their conclusions is
exactly what the flagship `quant_fit()` function automates. It executes all
three routes on the same data set and returns a single triangulated verdict,
with each route bootstrap-calibrated so the evidence is comparable:

* the **LC route** runs `select_model_ll()` on the raw person-by-item data
  (the chi-bar-squared likelihood-ratio bootstrap above);
* the **CC route** runs `cc_bootstrap_null()`, which calibrates the
  cancellation-check violation rate against a null simulated from the Rasch
  model fitted to the data, following Student & Read (2025);
* the **Kara route** runs `kara_bootstrap_null()`, which calibrates
  Karabatsos's global KL statistic against the same kind of Rasch bootstrap
  null, computed on an ability-banded matrix.

All three nulls are *marginal* parametric bootstraps — each replicate redraws
person abilities from the fitted Rasch latent distribution rather than reusing
fixed person estimates — and each returns a
`(1 + #{null >= observed}) / (B + 1)` percentile p-value. A route "supports
quantitative" when interval scaling is *not* rejected at the chosen null
percentile (default the 95th). The call fits and bootstraps all three routes,
so it is shown but not evaluated here:

```{r quant-fit, eval = FALSE}
v <- quant_fit(responses, mc.cores = 4, seed = 1)
v
#>
#> Quantitative structure - triangulated judgement
#> ===============================================
#> QUANTITATIVE (well supported) - a quantitative model is selected and
#> the conjoint axioms are satisfied; the routes converge.
#>
#> Routes supporting a quantitative reading: 3 of 3
#>
#> [LC]   Latent-structure model selection (raw data)
#>        selected RM (3 classes) - quantitative   [ yes ]
#> [CC]   Cancellation checks vs Rasch bootstrap null (Student & Read)
#>        double: rate 0.031, 44%ile of null (p = 0.564); triple 39%ile (p = 0.624)   [ yes ]
#> [Kara] Karabatsos KL vs Rasch bootstrap null (banded)
#>        global KL 0.006, 52%ile of null (p = 0.485)   [ yes ]
#>        per-cell KL median 0.004, Q3 0.009, 90% 0.014, max 0.021
#>
#> All three routes are bootstrap-calibrated (reject > 95%ile of the null):
#> LC by a chi-bar-squared LR bootstrap, CC and Kara by Rasch bootstrap nulls.
#> Note: N = 800; the axiom procedures are under-powered below ~1000 (Student & Read 2025).
#> Axiom banding (Kara) is unidimensional, so it can miss multidimensional
#> departures - read a low violation rate as within-scale additivity.
```

Either bootstrap-null route can also be run on its own — for example
`cc_bootstrap_null(responses, check = "double", B = 200, mc.cores = 4)` returns
an object carrying `observed`, `null`, `percentile`, `p_value`, and `reject`,
with `print` and `plot` methods — but `quant_fit()` is the recommended one-call
entry point when you want the triangulated verdict.

## Polytomous data

Everything above works unchanged on **polytomous** items — responses scored with
ordered integer categories `0, 1, ..., m`. QuantFit detects the scoring
automatically and fits each model natively, with no dichotomising and no loss of
information:

```{r polytomous, eval = FALSE}
# simulate partial-credit data: 6 items, 4 categories (0-3)
set.seed(1)
nP <- 1500; J <- 6; M <- 3
theta <- rnorm(nP)
delta <- matrix(seq(-1, 1, length.out = J), J, M) +
  matrix(seq(-0.8, 0.8, length.out = M), J, M, byrow = TRUE)
poly <- matrix(0L, nP, J)
for (j in seq_len(J)) {
  num <- cbind(0, t(apply(outer(theta, delta[j, ], "-"), 1, cumsum)))
  P <- exp(num) / rowSums(exp(num))
  poly[, j] <- rowSums(runif(nP) > t(apply(P, 1, cumsum)))
}

# the same calls as before — the six models fit the polytomous data directly
select_model_ll(poly, n_classes = 2, B = 99, seed = 1)
#> Selected model: RM
#> ... QUANTITATIVE (continuous: Rasch model)
```

Under the hood the latent-structure models carry a full category-probability
vector for each item and class; class monotonicity becomes stochastic ordering
of those category distributions across classes, invariant item ordering becomes
non-crossing expected item-score curves, and the quantitative models use the
partial credit model (a latent-class partial credit model for `LCR`, and a
marginal-maximum-likelihood partial credit model for `RM`). The conjoint routes
(`quant_fit()`'s CC and Kara components) recode each polytomous item into its
adjacent-category dichotomisations and condition on the total score before
running the cancellation checks. From the user's side, nothing changes: pass the
polytomous responses to `quant_fit()`, `select_model_ll()`, `fit_rm()`, and the
rest exactly as for dichotomous data.

## Which approach when?

* The **cancellation checks** and the **Karabatsos test** are axiom-focused:
  they ask directly whether additive-conjoint structure holds, without
  committing to a parametric latent variable model. They are the natural tools
  when the question is specifically about *measurement* in the
  representational sense.

* **Latent structure model selection** is model-focused and finer-grained: it
  distinguishes classificatory, ordinal, and quantitative structure and
  returns a fitted model you can interpret and report. `select_model_ll()` is
  the recommended selector; `compare_models()` is useful for a quick
  information-criterion overview, keeping its equal-parameter-count caveat in
  mind.

Running all three and comparing their conclusions is the most defensible
strategy: agreement across such different methods is strong evidence about the
structure of the variable, and disagreement is itself informative.

## References

Domingue, B. (2014). Evaluating the equal-interval hypothesis with test score
scales. *Psychometrika*, 79(1), 1-19.

Karabatsos, G. (2018). On Bayesian testing of additive conjoint measurement
axioms using synthetic likelihood. *Psychometrika*, 83(2), 321-332.

Torres Irribarra, D., & Diakow, R. Categorization, Ordering and Quantification:
Selecting a Latent Variable Model by Comparing Latent Structures.
