---
title: "Checking quantitative structure with QuantFit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking quantitative structure with QuantFit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1)
```

## The question

Before treating a psychological or educational test score as a *measurement*,
it is worth asking whether the data actually support a quantitative
interpretation of the latent variable at all. The differences between people
might be quantitative (a matter of *how much*), merely ordinal (a matter of
*rank*), or only classificatory (a matter of *kind*). QuantFit provides three
complementary, empirically testable answers to that question:

1. **Conjoint cancellation checks** ask whether the response probabilities
   obey the cancellation axioms of additive conjoint measurement — the
   order-theoretic conditions that must hold if persons and items sit on a
   common additive scale.

2. **The Karabatsos omnibus test** poses a closely related question in a
   Bayesian synthetic-likelihood framework, quantifying departures from
   additivity cell by cell.

3. **Latent structure model selection** fits a hierarchy of six latent
   variable models and asks which level of structure — classificatory,
   ordinal, or quantitative — the data actually warrant.

The three approaches come at the same underlying question from different
directions; used together they give a much fuller picture than any one alone.

```{r setup}
library(QuantFit)
```

We will use a single simulated data set throughout. It follows a Rasch model,
so all three approaches *should* conclude that the structure is quantitative.

```{r simulate}
set.seed(1)
n_persons <- 800
n_items   <- 12
theta <- rnorm(n_persons)                       # person abilities
beta  <- seq(-2, 2, length.out = n_items)       # item difficulties
prob  <- plogis(outer(theta, beta, "-"))
responses <- matrix(rbinom(n_persons * n_items, 1, prob),
                    n_persons, n_items)
dim(responses)
```

## 1. Conjoint cancellation checks

The cancellation checks operate on a matrix of counts cross-classifying
*sum score* against *item*. `PrepareChecks()` builds the required success
counts `n` and attempt counts `N` from a raw 0/1 response matrix.

```{r prepare}
prep <- PrepareChecks(responses)
dim(prep$N)
```

`ConjointChecks()` then samples submatrices, runs a constrained MCMC sampler
on each, and reports the proportion of cells whose observed proportion falls
outside the posterior credibility region implied by the axiom. Low violation
rates are consistent with additive (quantitative) structure. The sampler is
the computational bottleneck, so the call below is shown but not evaluated in
this vignette:

```{r conjoint, eval = FALSE}
cc <- ConjointChecks(prep$N, prep$n, n.mat = 100, check = "double",
                     mc.cores = 1)
cc@means            # unweighted and item-count-weighted violation rates
#> $unweighted
#> [1] 0.03
#>
#> $weighted
#> [1] 0.02
```

`check = "single"`, `"double"`, and `"triple"` test progressively stronger
cancellation axioms; `HiConjointChecks()` runs the full hierarchical test over
every 3x3 submatrix embedded in each 4x4 block. The bundled `rasch1000` object
is a pre-computed result you can explore immediately:

```{r rasch1000}
data(rasch1000)
summary(rasch1000)
```

## 2. The Karabatsos omnibus test

`KaraChecks()` implements Karabatsos (2018): it fits a two-stage isotonic
model to the observed proportions, simulates synthetic data sets under a
Beta-Binomial null, and uses a synthetic-likelihood importance sampler to
obtain a posterior for each cell. Departures from additivity are summarised
by a Kullback-Leibler divergence per cell (flagged when KL > 0.01) and a
global KL. This is a Monte Carlo procedure; realistic runs use `S` in the tens
of thousands, so it too is shown rather than evaluated:

```{r kara, eval = FALSE}
kc <- KaraChecks(prep$N, prep$n, S = 30000, N_synth = 100, mc.cores = 1,
                 verbose = FALSE)
kc$global_KL      # overall departure from additivity
kc$n_violations   # number of cells with KL > 0.01
kc$KL             # per-cell KL divergence matrix
```

For Rasch-generated data the global KL is small and few cells are flagged;
strongly non-additive data (for example 2PL data with dispersed
discriminations) produce many more violations. This R/C++ implementation is
validated against the original MATLAB `ACMtest.m` to a cell-level correlation
above 0.99.

## 3. Latent structure model selection

The third approach fits six nested-in-structure latent variable models:

| Model | Structure | Interpretation |
|-------|-----------|----------------|
| `UN`  | Unconstrained latent classes | Classificatory |
| `MON` | Class monotonicity | Ordinal |
| `IIO` | Invariant item ordering | Ordinal |
| `DM`  | Double monotonicity (MON + IIO) | Ordinal |
| `LCR` | Latent class Rasch | Quantitative (discrete) |
| `RM`  | Rasch model | Quantitative (continuous) |

You can fit any single model directly:

```{r fit-un}
un <- fit_un(responses, n_classes = 3, verbose = FALSE)
un
```

### Comparing all six by information criteria

`compare_models()` fits the whole set and ranks them. Note the caveat it is
built to expose: UN, MON, IIO, and DM all have the *same* number of
parameters, so BIC (which only rewards parsimony through parameter count)
cannot separate them — it will tend to pick UN or, for quantitative data, RM.

```{r compare}
cmp <- compare_models(responses, n_classes = 3, verbose = FALSE)
cmp
```

### Calibrated selection with the likelihood-ratio bootstrap

`select_model_ll()` addresses that limitation. It tests each constrained model
against its parent with a parametric bootstrap of the likelihood-ratio
statistic — the correct calibration when the models are nested by inequality
constraints and share a parameter count (the asymptotic null is
chi-bar-squared, not chi-squared). It walks down the hierarchy and uses BIC
only for the final RM-versus-LCR step, where the parameter counts really do
differ. Because it refits models on bootstrap samples it is moderately
expensive; here we use a small `B` for illustration:

```{r select-ll, eval = FALSE}
sel <- select_model_ll(responses, n_classes = 3, B = 99, seed = 1)
sel
#> Selected model: RM
#> Interpretation: QUANTITATIVE (continuous: Rasch model)
#>
#>   comparison  statistic p_value                 decision
#>     DM vs UN  204.38      0.01   rejected (p <= alpha)
#>    IIO vs UN  204.03      0.01   rejected (p <= alpha)
#>    MON vs UN    0.47      0.37   adequate (p > alpha)
```

A purely diagnostic alternative, `select_model_constraint()`, inspects whether
the unconstrained estimates *satisfy* each ordinal constraint (within a
tolerance) rather than testing them, and is much faster:

```{r select-constraint}
sc <- select_model_constraint(responses, n_classes = 3, seed = 1)
sc$selected
sc$interpretation
```

### Standard errors

`compute_se()` returns standard errors for a fitted model, either from the
observed information (default) or by nonparametric bootstrap. For a model
whose fitted estimates sit on an active monotonicity constraint, the
Hessian-based standard errors for the affected parameters are returned as `NA`
with a warning, and the bootstrap method should be used instead.

```{r se}
se <- compute_se(un, method = "hessian")
se
```

## Which approach when?

* The **cancellation checks** and the **Karabatsos test** are axiom-focused:
  they ask directly whether additive-conjoint structure holds, without
  committing to a parametric latent variable model. They are the natural tools
  when the question is specifically about *measurement* in the
  representational sense.

* **Latent structure model selection** is model-focused and finer-grained: it
  distinguishes classificatory, ordinal, and quantitative structure and
  returns a fitted model you can interpret and report. `select_model_ll()` is
  the recommended selector; `compare_models()` is useful for a quick
  information-criterion overview, keeping its equal-parameter-count caveat in
  mind.

Running all three and comparing their conclusions is the most defensible
strategy: agreement across such different methods is strong evidence about the
structure of the variable, and disagreement is itself informative.

## References

Domingue, B. (2014). Evaluating the equal-interval hypothesis with test score
scales. *Psychometrika*, 79(1), 1-19.

Karabatsos, G. (2018). On Bayesian testing of additive conjoint measurement
axioms using synthetic likelihood. *Psychometrika*, 83(2), 321-332.

Torres Irribarra, D., & Diakow, R. Categorization, Ordering and Quantification:
Selecting a Latent Variable Model by Comparing Latent Structures.
